Data exploration, quality control and statistical analyses of ChIP-exo experiments\vspace*{\fill}

### 2015-11-27 rmarkdown/prelims_paper.Rnw
Second version of the paper finished

% Recently, ChIP-exo has been developed to investigate protein-DNA
% interaction in higher resolution compared to popularly used
% ChIP-Seq. Although ChIP-exo has drawn much attention and is considered
% as powerful assay, currently no systematic studies have yet been
% conducted to determine which biases impact in a ChIP-exo's dataset
% quality.

@@

@@

While the number of produced ChIP-exo data keeps increasing,
characteristics of ChIP-exo data are not fully investigated
yet. First, DNA libraries generated by the ChIP-exo protocol seem to
be less complex than the libraries generated by ChIP-Seq (Mahony et
al., 2015 \cite{exo_review}), i.e. the number of positions to which
the reads can be aligned has been reduced due to the exonuclease
digestion. Second, although the are roughly the same amount of reads
in both strands, locally there may be more reads in one strand than in
the other. To address this challenges, we suggest a collection of
quality control visualizations to understand which of this biases are
present in ChIP-exo data and globally assess the enrichment and
library complexity of a ChIP-exo sample.We gathered ChIP-exo data from
diverse organisms: CTCF factor in human \cite{exo1}; ER factor in
human and FoxA1 factor in mouse (Serandour et al., 2013
\cite{exoillumina}); and generated $\sig$ factor in Escherichia Coli
(E. Coli) under aerobic ($ + O_2$) condition, and treated by
rifampicin by 0 and 20 minutes (courtesy of Professor Robert Landick's
lab).
% While the number of produced ChIP-exo data keeps increasing,
% characteristics of ChIP-exo data and its analysis are not fully
% investigated yet, including issues of sequence biases inherent to
% ChIP-exo data, choice of optimal statistical methods, and
% determination of optimal sequencing depth. However, currently the
% number of available ChIP-exo data is still limited and their
% sequencing depths are still insufficient for such investigation.

@@

Currently, research on statistical methods for ChIP-exo data is still
in its very early stage. Although many methods have been proposed to
identify protein binding sites from ChIP-Seq data (reviewed by
Wilbanks and Facciotti, 2012 \cite{evaluation} and Pepke and Wold,
2009 \cite{computation}), such as MACS (Zhang et al., 2008
\cite{macs}), CisGenome (Ji et al., 2008 \cite{cisgenome}) and
MOSAiCS (Kuan et al., 2009 \cite{mosaics}), these approaches reveal
protein binding sites in lower resolution, i.e., at an interval of
hundreds to thousands of base pairs. Furthermore, they report only one
``mode'' or ``predicted binding location'' per peak. Hence, these
methods are not appropriate to evaluate the potential of ChIP-exo data
for high resolution identification of protein binding sites. More
recently, deconvolution algorithms such as Deconvolution (Lun et al.,
2009 \cite{csdeconv}), GEM (Guo et al., 2012 \cite{gem}, an improved
version of Guo et al., 2010 \cite{gps} ) and PICS (Zhang et al., 2010
\cite{pics}) have been proposed to identify binding sites in higher
resolution using ChIP-Seq data. However, most of them are still not
tailored for ChIP-exo and PET and SET ChIP-Seq data in a unified
framework and as a result, currently available methods are not
appropriate for fair comparison between ChIP-exo and ChIP-Seq. To
address these limitations, we developed an improved of dPeak (Chung et
al., 2013 \cite{dpeak}), a high resolution binding site identification
(deconvolution) algorithm that we previously developed for PET and SET
ChIP-Seq data, so that it can also handle ChIP-exo data. The dPeak
algorithm implements a probabilistic model that accurately describes
the ChIP-exo and ChIP-Seq data generation process.

@@

\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics{../figs/diagrams/ChIPexo_diagram.png}
\caption{ChIP-exo diagram. The only difference with ChIP-Seq is the
  second step, where an exonuclease enzyme is added and it digest the
  DNA fragments starting from the 5' end until it finds a crosslink
  protein (this figure is taken from Furey, 2012 \cite{chipbeyond}).}
\label{fig:exo}
\end{wrapfigure}

@@

ChIP-exo experiments first capture millions of DNA fragments
(150 - 250 bp in length) that the protein under study interacts with
using random fragmentation of DNA and a protein-specific
antibody. Then, exonuclease is introduced to trim 5' end of each DNA
fragment to a fixed distance from the bound protein. As a result,
boundaries around the protein of interest constructed with 5' ends of
fragments are located much closed to bound protein compared to
ChIP-Seq. This step is unique to ChIP-exo and could potentially
provide significantly higher spatial resolution compared to
ChIP-Seq. Finally, high throughput sequencing of a small region (25 to
100 bp) at 5' end of each fragment generates millions of reads or
tags.

Since this protocol is a modification to ChIP-Seq, some of its
characteristics are still maintained in ChIP-exo while several new are
being found. Rhee and Pugh, 2011 \cite{exo1}, showed that ChIP-exo
generated fragments are located around the binding events more tightly
than ChIP-Seq generated fragments. Hence, it is of utmost importance
to understand how this new protocol works, specifically we focused on
understanding which biases are maintained from ChIP-Seq to ChIP-exo,
which biases are specific to ChIP-exo, how the current ChIP-Seq QC
guidelines behave in ChIP-exo and we developed a new QC pipeline
specific to ChIP-exo. A typical example is shown in figure
\ref{fig:exo_example}, which shows the $\sig$ and $\beta'_f$ factors
(and the $\beta$ factor only on top) on a E. Coli genome's region:
First, both regions show that the $\sig$ sample is enriched; second
the $\sig$ sample in the ChIP-exo panel shows that its respective form
in the ChIP-Seq panel may be form of more binding events; and third,
either of the $\beta$ factors show that the complexity of the ChIP-exo
libraries may be lower in some cases, i.e. the peak is mapped with
fewer positions.

\begin{figure}[H]
  \centering
  \includegraphics[width = .8 \textwidth,page = 107]{/p/keles/ChIPexo/volume5/OpenClosedComplexes/figs/peaks/ChIPseq_ChIPexo_peaks_rif20.pdf}
  \caption{An example of a PET ChIP-Seq peak compared to
    ChIP-exo. $\sig$, $\beta$ and $\beta_f'$ are subunits of the
    transcription initiation complex in \emph{E. Coli}. Gene regions
    on the reverse strand are highlighted in light blue and the region
    was centered respect to ChIP-Seq $\sig$ summit. The library
    complexity is lower in ChIP-exo than in ChIP-Seq but the
    resolution is increased, this can be observed at the $\sig$ peak
    that for ChIP-exo was created as the combination of at least three
    ChIP-exo ``sub-peaks''.}
  \label{fig:exo_example}
\end{figure}

@@

To our knowledge it doesn't exists a QC pipeline for ChIP-exo. Hence,
we analyzed how the current ChIP-Seq QC measures behave on ChIP-exo
data.

Table \ref{tab:qcbase} contains measures that are commonly used with
ChIP-Seq data. PBC stands for PCR bottleneck amplification and is
defined as the ratio between the number of positions to which exactly
one mapping read maps divided by the number of position to which at
least one mapping reads maps, the ENCODE project proposed a series of
ranges to interpret this quantity (which can be found in
\url{https://www.encodeproject.org/data-standards/2012-quality-metrics/}),
but it doesn't apply to ChIP-exo, if it did all the samples with
$\mbox{PBC} < 0.5$ should be reject due to be showing severe
bottlenecking. Planet et al., 2011 proposed the Standardized Standard
Deviation which defined as:

\[
\mbox{SSD} = \mbox{SD}_N / \sqrt{N} 
\]

where $\mbox{SD}_N$ is the standard deviation of the coverage of the
sample and $N$ is the number of reads. This measure is smaller for
ChIP-exo since the reads are located more tightly around binding
sites, but it may be biased since several fragments are being
digested by exonuclease enzyme, hence there are several position in
the genome where no fragment is being mapped. Finally the normalized
strand cross-correlation is defined as:

@@

where $y(\delta)$ is the SCC for a strand shift $\delta$, $r$ is the
Pearson correlation, $w_c$ is the proportion of tags mapped to
chromosome $c$ and $n_c^S$ is the tag count vector for strand $S$ and
chromosome $c$.

For ChIP-Seq data, the SCC will have two peaks: One when $\delta$ is
equal to reads length and one when $\delta$ is equal to the unobserved
fragment length, ideally the SCC function is maximized at this
point. Landt et al., 2012 \cite{encode_qc} explain that a successful if
this happens, it is marginally successful if its maximized at the read
length peak but still shows the fragment length summit and failed if
is maximized at the read length peak but the second summit is not
present.

\begin{figure}[H]
  \centering  
\includegraphics[width = .7\textwidth]{../figs/for_paper/scc_ctcf.pdf}
  \caption{SCC for CTCF factor in HeLa cell line for ChIP-exo (obtained from \cite{exo1}) and SET ChIP-Seq (obtained from \cite{encode1}) }
  \label{fig:scc}
\end{figure}

Figure \ref{fig:scc} shows the SCC function for both a ChIP-exo (in
red) and a SET ChIP-Seq (in green) for CTCF in HeLa cell line. The
ChIP-Seq dataset behaves as the successful case from \cite{encode_qc},
but the ChIP-exo doesn't. Additionally, since the exonuclease enzyme
digests the DNA fragments, then the shifts where the SCC may reach a
local or a global maximum are confounded. Hence, new QC measures are
necessary for ChIP-exo data.

\begin{table}[H]
  \centering
  \caption{Usual quality control indicators applied to the $\sigma^{70}$ samples. PBC stands PCR-bottleneck coefficient, SSD for standardized standard deviation, NSC for normalized strand cross-correlation coefficient (Another measure related to the strand cross-correlation curve is the RSC was omitted since input sample usually don't exists for ChIP-exo). $\sig$ samples were given by Robert Landick, FoxA1 and ER samples are from Serandour et al., 2013 \cite{exoillumina} and CTCF sample is from Rhee and Pugh, 2011 \cite{exo1}.}

@@

Chung et al., 2013 proposed the dPeak model to identify the
transcription factor binding sites in high resolution. Two models are
proposed: One to deconvolve the signal of single end tags (SET) data,
i.e. when only one end of the DNA fragment is sequenced; and another
to deconvolve the signal of paired end tags (PET) data, i.e. when both
ends of the DNA fragment are sequenced. For ChIP-exo, both ends are
sequenced but only the 5' end is considered, since is the side where
the exonuclease enzyme digests the DNA.

Given a peak region with $m$ positions, with $n$ DNA fragments and that
the region is generated by a known $g^*$ amount of binding
events. Each fragment is described by the start position where the
starting end is aligned and the strand from which the fragment is
sequenced. Since only one end is sequenced, then the fragment length
is not observed. dPeak assumes that the strand $D_i$ follows a
Bernoulli distribution with known parameter $p_D$, and the start
position $R_i$ for the $i$-th fragment is generated according to the
following procedure:

\begin{itemize}
\item If the fragment was sequenced from the forward strand ($D_i =
  1$):
  \begin{itemize}
  \item The reads belongs to the background: $R_i | Z_i = 0, D_i = 1
    \sim \mbox{Unif}(1 - \beta + 1, m)$
  \item The read belong to the $g$-th binding event: $R_i | Z_i = g,
    D_i = 1 \sim \mbox{N}(\mu_g - \delta , \sigma^2)$
  \end{itemize}
\item If the fragment was sequenced from the backward strand ($D_i =
  0$):
  \begin{itemize}
  \item The reads belongs to the background: $R_i | Z_i = 0, D_i = 1
    \sim \mbox{Unif}(1 , m + \beta  - 1)$
  \item The read belong to the $g$-th binding event: $R_i | Z_i = g,
    D_i = 1 \sim \mbox{N}(\mu_g + \delta , \sigma^2)$
  \end{itemize}

@@

and the binding event to which the $i$-th fragment belongs $Z_i = g$
is generated from a

$\mbox{Multinomial}(\pi_0,\pi_1,\cdots,\pi_{g^*})$.

@@

Chung et al., 2013 \cite{dpeak} derived an initialization strategy for
the stochastic EM algorithm where each region is initialized by
separate independently and the binding sites locations $\mu_g,
g=1,\cdots,g^*$ are initialized by a uniform partition.

For this work, we considered a new strategy where the all the regions
are ranked by a statistical measure (as the $-\log(p)$ where $p$ is
the empirical p.value of the test $H_0: \mathcal{R} \text{ is
  enriched}$, where $\mathcal{R}$ is a genomic region) or a empirical
quantity (as the number of ChIP tag counts in the region) and use as
initial estimates of the $\delta$ and $\sigma$ parameters the average
of the estimates from the highest ranked $N_{\text{top}}$ peaks.

@@

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of \Sexpr{bin_size} non-overlapping
window after extending reads by \Sexpr{frag_len} to their 3' end
directions. ChIP tag counts in ChIP-exo data were linearly related to
ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
potential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that background distribution of ChIP-exo data is less
homogeneous than that of ChIP-Seq data.

\begin{figure}[H]
  \centering
  \includegraphics[width = .46\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  \includegraphics[width = .46\textwidth]{../figs/for_paper/forward_strand_ratio_comp_old.pdf}
\includegraphics[width = .46\textwidth,page = 1]{../figs/for_paper/eukaryotic_bias_CTCF.pdf}
  \includegraphics[width = .46\textwidth,page = 2]{../figs/for_paper/eukaryotic_bias_CTCF.pdf}
  \caption{ A) ChIP signal in ChIP-exo was linearly related to that in
    PET ChIP-Seq data in the region with high ChIP tag counts. In
    contrast, there were clear differences in their background
    distribution, where several ChIP-exo background regions were
    empty. B) In ChIP-exo data, strands of reads were significantly
    less balanced in the regions with potential binding sites compared
    to SET ChIP-Seq data. C) ChIP tag counts increase linearly as
    mappability scores increases. D) ChIP tag counts increase linearly
    as GC content score increases when GC content is less than 0.6 and
    then ChIP tag counts decrease as GC content increases. }
  \label{fig:comp}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/for_paper/coverage_diagram2.pdf}
  \caption{Diagram of the ChIP-exo Quality control pipeline. The
    genome is partitioned into several regions by removing the empty
    segments and for each regions we calculate a collection of QC
    indicators, then for each region $R_k$ a vector of summary
    statistics $T(R_k)$ is calculated.}
  \label{fig:qcdiagram}
\end{figure}

@@

We considered the creation of a novel QC measure for ChIP-exo called
local normalized SCC or $\mbox{local-NSC}$. This measure

@@

where $y(\delta)$ is the observed SCC for shift $x_\delta$,
$\epsilon_\delta \sim N(0,\sigma_f^2)$. Finally the $\mbox{local-NSC}$
is defined as a signal-to-noise ratio by using:

\begin{align}
    \mbox{local-NSC} &= \frac{max_{x_\delta} \hat{f}(x_\delta)}{\hat{\sigma}_f} \label{locNSC}
\end{align}

\begin{figure}[H]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/for_paper/local_SCC_example.pdf}
  \caption{Using the FoxA1 sample from \cite{exoillumina}. A) shows an
    example region for the 3 replicates. The library complexity seem
    to be comparable between the two top rows and both outperform the
    last row. The number of reads for the first replicate is higher
    than the number of reads for the other replicates. B) Shows the
    local SCC for the 3 replicates, this measure is comparable among
    the 3 replicates. The first replicate shows the lowest residual
    error, while the third one shows the highest. In blue, a ``loess''
    regression model is fitted.}
\label{fig:locSCC}
\end{figure}

@@

Using the mouse-FoxA1 experiment from \cite{exoillumina} and the
relationship between this two quantities, library complexity and
sample enrichment was explored. Figure \ref{fig:enrich}A shows hexbin
plots illustrating the interaction between these two
quantities. Notice how there are two strong arms in each panel: The
first one corresponds to regions with low $\mbox{ARC}$ values and
varying $\mbox{URCR}$ values across the $(0,1]$ interval, while the
second one shows a decreasing trend in $\mbox{URCR}$ as $\mbox{ARC}$
increases. When an experiment shows a higher degree of enrichment,
then the separation of this two arms is more noticeable, since the
second arm corresponds to possibly enriched regions
(\ref{fig:enrich2}A and \ref{fig:enrich2}B).

>>>>

ChIP-exo is a modification of the ChIP-seq protocol for high
resolution mapping of transcription factor binding sites. Although
many aspect of the ChIP-exo data analysis are similar to those of
ChIP-Seq, ChIP-exo presents a number of unique challenges. We present
a quality control pipeline that analyzes ChIP-exo's strand imbalance
and library complexity. Assessment of these biases and artifacts are
facilitated through diagnostic plots and summary statistical that
compare a large portion of the genome as partitioned into island with
high depth regions.

We systematically evaluated diverse aspects of ChIP-exo and found the
following characteristics: First, ChIP-exo's background is quite
different than ChIP-Seq's. Second, although often assumed in ChIP-exo
data analysis methods, the ``peak pair'' assumption doesn't hold
locally in real ChIP-exo data. Third, we compared Paired End (PE)
ChIP-Seq with ChIP-exo and found that both protocols are comparable in
resolution and sensitivity for closely located binding event, but as
the distance between binding event increases ChIP-exo shows higher
sensitivity than PE ChIP-Seq. Finally, for given fixed sequencing
depth, ChIP-exo provides higher sensitivity, specificity and spatial
resolution than PE ChIP-Seq.

@@

% We also systematically compare multiple ChIP-seq and ChIP-exo
% datasets to quantify differences between these two protocols. Our
% analysis indicate that spatial resolution of ChIP-exo is comparable to
% that of paired-end ChIP-seq and both of them are significantly better
% than resolution of single-end ChIP-seq. Furthermore, for a given fixed
% sequencing depth, ChIP-exo provides higher sensitivity, specificity,
% and spatial resolution than pairedend ChIP-seq.

@@

While the number of produced ChIP-exo data keeps increasing,
characteristics of ChIP-exo data are not fully investigated
yet. First, DNA libraries generated by the ChIP-exo protocol seem to
be less complex than the libraries generated by ChIP-Seq (Mahony et
al., 2015 \cite{exo_review}), i.e. the number of positions to which
the reads can be aligned has been reduced due to the exonuclease
digestion. Second, although the are roughly the same amount of reads
in both strands, locally there may be more reads in one strand than in
the other. To address this challenges, we suggest a collection of
quality control visualizations to understand which of this biases are
present in ChIP-exo data and globally assess the enrichment and
library complexity of a ChIP-exo sample.We gathered ChIP-exo data from
diverse organisms: CTCF factor in human \cite{exo1}; ER factor in
human and FoxA1 factor in mouse (Serandour et al., 2013
\cite{exoillumina}); and generated $\sig$ factor in Escherichia Coli
(\emph{E. Coli}) under aerobic ($ + O_2$) condition, and treated by
rifampicin by 0 and 20 minutes (courtesy of Professor Robert Landick's
lab).
% While the number of produced ChIP-exo data keeps increasing,
% characteristics of ChIP-exo data and its analysis are not fully
% investigated yet, including issues of sequence biases inherent to
% ChIP-exo data, choice of optimal statistical methods, and
% determination of optimal sequencing depth. However, currently the
% number of available ChIP-exo data is still limited and their
% sequencing depths are still insufficient for such investigation.

@@

Currently, research on statistical methods for ChIP-exo data is still
in its very early stage. Although many methods have been proposed to
identify protein binding sites from ChIP-Seq data (reviewed by
Wilbanks and Facciotti, 2012 \cite{evaluation} and Pepke and Wold,
2009 \cite{computation}), such as MACS (Zhang et al., 2008
\cite{macs}), CisGenome (Ji et al., 2008 \cite{cisgenome}) and
MOSAiCS (Kuan et al., 2009 \cite{mosaics}), these approaches reveal
protein binding sites in lower resolution, i.e., at an interval of
hundreds to thousands of base pairs. Furthermore, they report only one
``mode'' or ``predicted binding location'' per peak. Hence, these
methods are not appropriate to evaluate the potential of ChIP-exo data
for high resolution identification of protein binding sites. More
recently, deconvolution algorithms such as Deconvolution (Lun et al.,
2009 \cite{csdeconv}), GEM (Guo et al., 2012 \cite{gem}, an improved
version of Guo et al., 2010 \cite{gps} ) and PICS (Zhang et al., 2010
\cite{pics}) have been proposed to identify binding sites in higher
resolution using ChIP-Seq data. However, most of them are still not
tailored for ChIP-exo and PE and SE ChIP-Seq data in a unified
framework and as a result, currently available methods are not
appropriate for fair comparison between ChIP-exo and ChIP-Seq. To
address these limitations, we developed an improved of dPeak (Chung et
al., 2013 \cite{dpeak}), a high resolution binding site identification
(deconvolution) algorithm that we previously developed for PE and SE
ChIP-Seq data, so that it can also handle ChIP-exo data. The dPeak
algorithm implements a probabilistic model that accurately describes
the ChIP-exo and ChIP-Seq data generation process.

@@

\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics{../figs/diagrams/ChIPexo_diagram.png}
\caption{ChIP-exo diagram. The only difference with ChIP-Seq is the
  second step, where an exonuclease enzyme is added and it digest the
  DNA fragments starting from the 5' end until it finds a crosslink
  protein (this Figure is taken from Furey, 2012 \cite{chipbeyond}).}
\label{fig:exo}
\end{wrapfigure}

@@

ChIP-exo experiments first capture millions of DNA fragments (150 -
250 bp in length) that the protein under study interacts with using
random fragmentation of DNA and a protein-specific antibody. Then,
exonuclease is introduced to trim 5' end of each DNA fragment to a
fixed distance from the bound protein. As a result, boundaries around
the protein of interest constructed with 5' ends of reads are located
much closed to bound protein compared to ChIP-Seq. This step is unique
to ChIP-exo and could potentially provide significantly higher spatial
resolution compared to ChIP-Seq. Finally, high throughput sequencing
of a small region (25 to 100 bp) at 5' end of each fragment generates
millions of reads or tags.

Since this protocol is a modification to ChIP-Seq, some of its
characteristics are still maintained in ChIP-exo while several new are
being found. Rhee and Pugh, 2011 \cite{exo1}, showed that ChIP-exo
generated fragments are located around the binding events more tightly
than ChIP-Seq generated fragments. Hence, it is of utmost importance
to understand how this new protocol works, specifically we focused on
understanding which biases are maintained from ChIP-Seq to ChIP-exo,
which biases are specific to ChIP-exo, how the current ChIP-Seq QC
guidelines behave in ChIP-exo and we developed a new QC pipeline
specific to ChIP-exo. A typical example is shown in Figure
\ref{fig:exo_example}, which shows the $\sig$ and $\beta'_f$ factors
(and the $\beta$ factor only on top) on a \emph{E. Coli} genome's
region: First, both regions show that the $\sig$ sample is enriched;
second the $\sig$ sample in the ChIP-exo panel shows that its
respective form in the ChIP-Seq panel may be form of more binding
events; and third, either of the $\beta$ factors show that the
complexity of the ChIP-exo libraries may be lower in some cases,
i.e. the peak is mapped with fewer positions. In \emph{E. Coli}
multiple promoters may be located close to each other, hence we may
suspect that the ChIP-Seq peak is occasioned by several promoter and
in the ChIP-exo panel we can observe that such is the case.

\begin{figure}[H]
  \centering
  \includegraphics[width = .8 \textwidth,page = 107]{/p/keles/ChIPexo/volume5/OpenClosedComplexes/figs/peaks/ChIPseq_ChIPexo_peaks_rif20.pdf}
  \caption{An example of a PE ChIP-Seq peak compared to
    ChIP-exo. $\sig$, $\beta$ and $\beta_f'$ are subunits of the
    transcription initiation complex in \emph{E. Coli}. Gene regions
    on the reverse strand are highlighted in light blue and the region
    was centered respect to ChIP-Seq $\sig$ summit. The library
    complexity is lower in ChIP-exo than in ChIP-Seq but the
    resolution is increased, this can be observed at the $\sig$ peak
    that for ChIP-exo was created as the combination of at least three
    ChIP-exo ``sub-peaks''.}
  \label{fig:exo_example}
\end{figure}

@@

We start our exploration by investigating the suitability of the
current state of the art QC pipeline for ChIP-Seq to ChIP-exo.

Table \ref{tab:qcbase} contains measures that are commonly used with
ChIP-Seq data. PBC stands for PCR bottleneck amplification and is
defined as the ratio between the number of positions to which exactly
one mapping read maps divided by the number of position to which at
least one mapping reads maps, the ENCODE project proposed a series of
ranges to interpret this quantity (which can be found in
\url{https://www.encodeproject.org/data-standards/2012-quality-metrics/}),
but it doesn't apply to ChIP-exo, if it did all the samples with
$\mbox{PBC} < 0.5$ should be reject due to be showing severe
bottlenecking. The normalized strand cross-correlation (NSC) is
defined as:

@@

where $y(\delta)$ is the SCC for a strand shift $\delta$, $r$ is the
Pearson correlation, $w_c$ is the proportion of reads mapped to
chromosome $c$ and $n_c^S$ is the read count vector for strand $S$ and
chromosome $c$.

\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth]{../figs/for_paper/SCC.png}
  \caption{SCC interpretation for ChIP-Seq from \cite{encode_qc}. In
    all panels, we observe shift $\delta$ vs SCC. The quality of
    ChIP-Seq is determined by the relative heights of the ``phantom
    peak'' and the local maximum when the shift is equal to the
    fragment length.}
  \label{fig:scc_landt}
\end{figure}

For ChIP-Seq data, the SCC will have two peaks: One when $\delta$ is
equal to read length and one when $\delta$ is equal to the unobserved
fragment length. The local maximum at the read length may be
occasioned by the duplicated reads or unmappable regions
\cite{carroll.qc}, or simply by reads in both strand that are mapped
in the same position. Figure \ref{fig:scc_landt} shows how to
interpret the SCC, the quality of a ChIP-Seq sample is going to be
determined by the relative height of the ``phantom peak'' and the
local maximum at the fragment length.

\begin{figure}[H]
  \centering  
\includegraphics[width = .7\textwidth]{../figs/for_paper/scc_ctcf.pdf}
  \caption{SCC for CTCF factor in HeLa cell line for ChIP-exo (obtained from \cite{exo1}) and SE ChIP-Seq (obtained from \cite{encode1}) }
  \label{fig:scc}
\end{figure}

Figure \ref{fig:scc} shows the SCC function for both a ChIP-exo (in
red) and a SE ChIP-Seq (in green) for CTCF in HeLa cell line. The
ChIP-Seq dataset behaves as the successful case from \cite{encode_qc},
but the ChIP-exo doesn't. Additionally, since the exonuclease enzyme
digests the DNA fragments, then the shifts where the SCC may reach a
local or a global maximum are confounded. Hence, new QC measures are
necessary for ChIP-exo data.

\begin{table}[H]
  \centering
  \caption{Usual quality control indicators applied to the $\sigma^{70}$ samples. PBC stands PCR-bottleneck coefficient (0-0.5 is severe bottlenecking, 0.5-0.8 is moderate bottlenecking, 0.8-0.9 is mild bottlenecking, while 0.9-1.0 is no bottlenecking), NSC for normalized strand cross-correlation coefficient (Another measure related to the strand cross-correlation curve is the RSC was omitted since input sample usually don't exists for ChIP-exo). $\sig$ samples were given by Professor Robert Landick's Lab, FoxA1 and ER samples are from Serandour et al., 2013 \cite{exoillumina} and CTCF sample is from Rhee and Pugh, 2011 \cite{exo1}. }

@@

Chung et al., 2013 proposed the dPeak model to identify the
transcription factor binding sites in high resolution. Two models are
proposed: One to deconvolve the signal of single end reads (SE) data,
i.e. when only one end of the DNA fragment is sequenced; and another
to deconvolve the signal of paired end reads (PE) data, i.e. when both
ends of the DNA fragment are sequenced. For ChIP-exo, only the 5' end
is considered, since is the side where the exonuclease enzyme digests
the DNA.

\begin{figure}[H]
  \centering
  \includegraphics[width = .6\textwidth]{/p/keles/ChIPexo/volume3/ChIPexo/poster/dpeak.png}
  \caption{A schematic illustration of dPeak's model when the read is
    located around binding site $\mu_g$}
\end{figure}

Given a peak region with $m$ positions, with $n$ DNA reads and that
the region is generated by a known $g^*$ amount of binding
events. Each read is described by the start position where the 5' end
is aligned and the strand from which the fragment is sequenced. Since
only one end is sequenced, then the fragment length is not
observed. dPeak assumes that the strand $D_i$ follows a Bernoulli
distribution with known parameter $p_D$, the binding event is not
observed and it follows a
$\mbox{Multinomial}(\pi_0,\pi_1,\cdots,\pi_{g^*})$. And the 5' end
$R_i$ for the $i$-th fragment is generated according to the following
procedure:

\begin{itemize}
\item If the read was sequenced from the forward strand ($D_i = 1$):
  \begin{itemize}
  \item The read belongs to the background: $R_i | Z_i = 0, D_i = 1
    \sim \mbox{Unif}(1, m)$
  \item The read belong to the $g$-th binding event: $R_i | Z_i = g,
    D_i = 1 \sim \mbox{N}(\mu_g - \delta , \sigma^2)$
  \end{itemize}
\item If the read was sequenced from the backward strand ($D_i =
  0$):
  \begin{itemize}
  \item The read belongs to the background: $R_i | Z_i = 0, D_i = 1
    \sim \mbox{Unif}(1 , m)$
  \item The read belong to the $g$-th binding event: $R_i | Z_i = g,
    D_i = 1 \sim \mbox{N}(\mu_g + \delta , \sigma^2)$
  \end{itemize}

@@

@@

Chung et al., 2013 \cite{dpeak} derived an initialization strategy for
the stochastic EM algorithm where for each region, the binding sites
locations $\mu_g, g=1,\cdots,g^*$ were initialized equally spaced
around the region and the parameters $\delta$ and $\sigma$ were
estimated by separate for all regions. This is problematic since this
initialization strategy does not depend on the data, specially in
regions with low signal to noise ration and with a low amount of
reads.

For this work, we considered a new strategy where the $mu_g$'s are
initialized by partitioning the region into $g^*$ bins and finding the
coverage's local maxima in each bin and we initialize $\delta$ and
$\sigma$ by ranking the regions by the signal strength or an empirical
quantity (as the number of ChIP tag counts in the region), fit the
model for the top $N_{\text{top}}$ regions and then use as initial
values the average of the $\sigma$ and $\delta$ estimates for those
regions.

@@

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of \Sexpr{bin_size} non-overlapping
window after extending reads by \Sexpr{frag_len} to their 3' end
directions. ChIP tag counts in ChIP-exo data were linearly related to
ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
potential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that for ChIP-exo data a much smaller portion of the
genome is expected to be background.

\begin{figure}[H]
  \centering
  \includegraphics[width = .46\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  \includegraphics[width = .46\textwidth]{../figs/for_paper/forward_strand_ratio_comp_old.pdf}
\includegraphics[width = .46\textwidth,page = 1]{../figs/for_paper/eukaryotic_bias_CTCF.pdf}
  \includegraphics[width = .46\textwidth,page = 2]{../figs/for_paper/eukaryotic_bias_CTCF.pdf}
  \caption{ A) Hexbin plot of PE ChIP-Seq bin counts VS ChIP-exo bin
    counts. B) Forward Strand Ratio densities for SE ChIP-Seq and
    ChIP-exo peaks. C) ChIP tag counts increase linearly as
    mappability scores increases. D) ChIP tag counts increase linearly
    as GC content score increases when GC content is less than 0.6 and
    then ChIP tag counts decrease as GC content increases. }
  \label{fig:comp}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/for_paper/coverage_diagram2.pdf}
  \caption{Flowchart of the ChIP-exo Quality control pipeline.}
  % The genome is partitioned into several regions by removing the empty
  % segments and for each regions we calculate a collection of QC
  % indicators, then for each region $R_k$ a vector of summary
  % statistics $T(R_k)$ is calculated.
  \label{fig:qcdiagram}
\end{figure}

@@

We considered the creation of a novel QC measure for ChIP-exo called
local normalized SCC or $\mbox{local-NSC}$. This measure is built by
calculating the SCC for a given region:

@@

where $y(\delta)$ is the observed SCC for shift $x_\delta$,
$\epsilon_\delta \sim N(0,\sigma_f^2)$.

Finally the $\mbox{local-NSC}$ is defined as:

\begin{align}
    \mbox{local-NSC} &= \frac{\mbox{max}_{x_\delta} \hat{f}(x_\delta)}{\hat{\sigma}_f} 
\label{locNSC}
\end{align}

Local NSC gives a signal-to-noise ratio for a given region in a
ChIP-exo sample.

\begin{figure}[H]
  \centering
  \includegraphics[width = .9\textwidth]{../figs/for_paper/local_SCC_example.pdf}
  \caption{Using the FoxA1 sample from \cite{exoillumina}. A) shows an
    example region for the 3 replicates. The library complexity seem
    to be comparable between the two top rows and both outperform the
    last row. The number of reads for the first replicate is higher
    than the number of reads for the other replicates. B) Shows the
    local SCC for the 3 replicates, this measure is comparable among
    the 3 replicates. The first replicate shows the lowest residual
    error, while the third one shows the highest. In blue, a ``loess''
    regression model is fitted.}
\label{fig:locSCC}
\end{figure}

@@

Figure \ref{fig:enrich}A shows hexbin plots illustrating the
relationship between library complexity and sample enrichment for the
FoxA1 in mouse cells from \cite{exoillumina}. Notice how there are two
strong arms in each panel: The first one corresponds to regions with
low $\mbox{ARC}$ values and varying $\mbox{URCR}$ values across the
$(0,1]$ interval, while the second one shows a decreasing trend in
$\mbox{URCR}$ as $\mbox{ARC}$ increases. When an experiment shows a
higher degree of enrichment, then the separation of this two arms is
more noticeable, since as Figures \ref{fig:enrich2}A and
\ref{fig:enrich2}B show the ARC VS URCR plots but filtering out the
regions formed by a low number of unique positions (10 and 30
respectively), hence this second arm corresponds to possibly enriched
regions.
